---
title: "Introduction to netimpute"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to netimpute}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
set.seed(2024)
```

## Why impute both sides at once

Missing data in a network study rarely stays on one side. The respondent who
skipped half the questionnaire is often the same one whose nominations are
missing, so the attributes you would use to impute the ties are themselves
incomplete — and the ties you would use to impute the attributes are too.

`netimpute` runs both through a single chained-equations loop. Within each
iteration every incomplete attribute and every incomplete network is visited
exactly once, in one interleaved, randomly ordered sequence. Whatever was
updated most recently — attribute or network — feeds the next target's
predictors. Attributes are imputed from the other attributes plus node-level
network measures; ties are imputed from a dyad-level model built from the
attributes and the other networks.

```{r}
library(netimpute)
```

## The example data

Thirty employees, two networks, and four attributes. Missingness is scattered
across both sides.

```{r}
n <- 30

make_net <- function(p) {
  m <- matrix(rbinom(n * n, 1, p), n, n)
  diag(m) <- 0
  m
}
punch <- function(m, k) {
  m[sample(which(row(m) != col(m)), k)] <- NA
  m
}

friends <- punch(make_net(0.15), 60)
advice  <- punch(make_net(0.10), 40)

staff <- data.frame(
  age    = round(rnorm(n, 40, 9)),
  female = rbinom(n, 1, 0.5),
  unit   = sample(c("sales", "eng", "hr"), n, TRUE),
  perf   = rnorm(n)
)
staff$age[sample(n, 4)]  <- NA
staff$perf[sample(n, 5)] <- NA
staff$unit[sample(n, 3)] <- NA

nets <- list(friends = friends, advice = advice)

colSums(is.na(staff))
vapply(nets, function(m) sum(is.na(m)), integer(1))
```

## A first imputation

`netmice()` is the entry point. `m` is the number of imputed datasets,
`maxit` the number of iterations per chain — the same meaning they have in
`mice`.

```{r}
fit <- netmice(staff, nets, m = 5, maxit = 10, seed = 1, printFlag = FALSE)
fit
```

The message about inferred attribute types is worth reading: `netimpute`
decides whether each column is continuous, binary, or multinomial, and that
decision drives both how the column is imputed and what its predictor terms
are called. Override it with `attr_types` if it guesses wrong.

`complete_netmice()` extracts one completed dataset — attributes and networks
together.

```{r}
done <- complete_netmice(fit, 1)
str(done$data)
sum(is.na(done$net_list$friends))
```

## Checking convergence

`plot()` shows the usual chain diagnostics on three pages: the mean and
variance of the imputed values per iteration for each attribute; density,
reciprocity, transitivity, isolate count and average inverse geodesic distance
for each network; and, for each network that had missing ties, the mean and
variance of the *imputed ties only*. You are looking for the chains to mix, not
to trend.

Read the third page first. The network diagnostics on the second page are
computed on the whole completed network, observed ties included, so they are
diluted by the missingness fraction: with 7% of dyads missing, a swing of 0.20
in the density of the imputed cells shows up as a movement of only about 0.014
in the network's overall density. A flat trace on page two is therefore *not*
evidence of convergence when missingness is low — the imputed-tie traces are the
sensitive ones, and they are what the attribute page has always shown for
attributes.

```{r, fig.alt = "Convergence diagnostics across iterations"}
plot(fit, nets = "friends", ask = FALSE)
```

If the imputed-tie traces are still drifting at the last iteration, raise
`maxit`. A steadily climbing imputed-tie mean is the signature of tie inflation
specifically, rather than of slow mixing in general.

The numbers behind the plots are on the fitted object: `chainMean`/`chainVar`
for attributes, `netChain` for the whole-network diagnostics, and
`netImpMean`/`netImpVar` for the imputed ties.

## Using the completed data

There is no pooling helper: fit your model to each completed dataset and pool
with `mice::pool()`. Anything you derive from the imputed networks — degree,
centrality, a contagion term — is computed inside the loop, so its uncertainty
flows into the pooled standard errors.

```{r}
fits <- lapply(seq_len(fit$m), function(k) {
  d <- complete_netmice(fit, k)
  d$data$friends_indeg <- colSums(d$net_list$friends)
  lm(perf ~ age + friends_indeg, data = d$data)
})

summary(mice::pool(fits))
```

## Predictor names: the one thing to get right

If you want to steer the imputation models yourself, you need the names of the
internally generated terms — and the naming rule depends on the attribute
type. **Continuous** attributes get bare `_ego` / `_alter` / `_absdiff` terms.
**Binary and multinomial** attributes are dummy-coded, one term per
non-reference level.

Rather than guessing, list them. `dyad_regression()` builds exactly the design
`netmice()` uses at a network visit, and `fit = FALSE` stops before fitting:

```{r}
terms <- names(dyad_regression(nets, staff, target = "friends",
                               fit = FALSE)$data)
setdiff(terms, c("i", "j", "y"))
```

Note the difference between `age` (continuous) and `female` / `unit`
(categorical):

```{r}
grep("^age",    terms, value = TRUE)
grep("^female", terms, value = TRUE)
grep("^unit",   terms, value = TRUE)
```

`female` is 0/1, so it is inferred as **binary** and its term is
`female_ego_1` — not `female_ego`. Had it been coded `"m"`/`"f"`, the term
would be `female_ego_m` or `female_ego_f`, depending on which level sorts
first and becomes the reference. This is the most common source of "why
doesn't my `models` formula work".

The node-level names an *attribute* model can reference come from
`net_measures()`, prefixed with the network name when more than one network is
supplied:

```{r}
g <- igraph::graph_from_adjacency_matrix(
  replace(friends, is.na(friends), 0), mode = "directed", diag = FALSE)
head(names(net_measures(g, staff)), 12)
```

Both lists are reproduced in full in the "Dyad-level term names" and
"Implemented network measures" sections of `?netmice`.

## Steering the models

`models` protects a hypothesised relationship: the terms you name always enter
that target's imputation model, whatever the automatic predictor machinery
decides. Formulas are given as strings; the left-hand side names the target.

```{r}
fit2 <- netmice(
  staff, nets, m = 5, maxit = 10, seed = 1, printFlag = FALSE,
  models = list(
    "friends ~ female_ego_1:reciprocity",
    "perf    ~ age + friends_indegree"
  )
)
```

The first formula is an interaction between an ego attribute and reciprocity:
*do women reciprocate friendship nominations differently?* Under the default
tie-wise updater (`net_update = "gibbs"`) the reciprocity factor in that
interaction is re-read from the live network after **every single tie draw**,
not held at its value from the start of the visit — which is what makes an
endogenous interaction meaningful here.

## Leaner models

By default every attribute and every network feature is a candidate predictor,
with a PCA safeguard if that gets too wide for the number of observed cases.
`netquickpred()` instead screens predictors per target by correlation, adds
predictors-of-predictors recursively, and prunes collinear survivors.

```{r}
qp <- netquickpred(staff, nets)
qp
```

Pass it straight to `netmice()`, or use `predictor_selection = "quickpred"` to
have it computed internally:

```{r}
fit3 <- netmice(staff, nets, m = 5, maxit = 10, seed = 1, printFlag = FALSE,
                predictor_selection = qp)
```

## Constraints on where ties can exist

`structural` marks cells that are zero *by design* — nominations that could not
have been made, such as ties into another school class. They are fixed at zero,
excluded from the model, and never imputed.

```{r}
structural <- matrix(FALSE, n, n)
structural[1:10, 11:30] <- TRUE
structural[11:30, 1:10] <- TRUE

friends_split <- friends
friends_split[structural] <- 0

fit4 <- netmice(staff, list(friends = friends_split), structural = structural,
                m = 3, maxit = 5, seed = 1, printFlag = FALSE)
all(complete_netmice(fit4, 1)$net_list$friends[structural] == 0)
```

`net_dependence` expresses logical relations *between* networks: `necessary`
means a tie in the parent is required for a tie in the child, `forbidden` that
two ties are mutually exclusive. Unlike `structural`, these are re-evaluated
against the partner network's current state at every visit, so the constrained
set adapts as the partner is itself imputed.

```{r}
# advice can only run along friendship ties
friends_full <- make_net(0.25)
advice_sub   <- friends_full * matrix(rbinom(n * n, 1, 0.5), n, n)

fit5 <- netmice(
  staff,
  list(friends = punch(friends_full, 50), advice = punch(advice_sub, 50)),
  net_dependence = list(necessary = list(c("friends", "advice"))),
  m = 3, maxit = 5, seed = 1, printFlag = FALSE
)

out <- complete_netmice(fit5, 1)$net_list
all(out$advice[out$friends == 0 & row(out$friends) != col(out$friends)] == 0)
```

## Directed and undirected networks

A network counts as undirected when the matrix you supply is symmetric in its
values **and** in its pattern of missing cells — exactly the case in which
cells $(i,j)$ and $(j,i)$ record one tie rather than two. That classification
is made once, from the data you pass in, and an undirected network is imputed
one tie at a time, so it comes back undirected.

```{r}
und <- make_net(0.25)
und[lower.tri(und)] <- t(und)[lower.tri(und)]
und[sample(which(upper.tri(und)), 40)] <- NA
und[lower.tri(und)] <- t(und)[lower.tri(und)]

fit6 <- netmice(staff, list(contact = und), m = 3, maxit = 5, seed = 1,
                printFlag = FALSE)
isSymmetric(unname(complete_netmice(fit6, 1)$net_list$contact))
```

The classification cannot be recovered after the fact: a directed network whose
asymmetry happens to sit entirely in its missing cells becomes symmetric the
moment those are filled. Supply the matrix with its `NA`s intact and the
distinction is preserved.

## Where to go next

- `?netmice` — the full argument reference, plus complete lists of dyad-level
  term names and implemented network measures.
- `?netquickpred` — how predictors are screened, and the tuning arguments.
- `?dyad_regression` — the tie model on its own, usable as a standalone
  MR-QAP-style regression.
- `?net_measures`, `?net_predictors` — the node-level measures.

The methodology is described in Krause (2026),
[arXiv:2608.05895](https://doi.org/10.48550/arXiv.2608.05895).
